Debugging by \textit{lastChange}

### 2010-12-08 papers/ecoop2011/ecoop2011-lastChange.tex
re-edit pass complete

\begin{abstract}
Developers often seek the origins of wrong values they see in their
debugger. Their search must be backwards in time: the code causing the
wrong value executed before the wrong value appeared. Searching with
breakpoint- or log- based debuggers demands persistence and
significant experience with the application being debugged. We propose a new 
practical technique called \textit{lastChange}, which automatically
locates the last point that a variable or an object property was changed and provides related
information such as the call stack and program state. Starting from a
program halted on a breakpoint, the \textit{lastChange} solution
applies queries to the live program during re-execution. Thus
\textit{lastChange} combines the flexibility of breakpoint debugging
with the expressive power of log-based query debugging.  Contrary to
other replay-based approaches which require exactly the same
re-executions (fully deterministic), \textit{lastChange} only requires \textit{bug 
reproducibility}, a test case is available which reproduces the bug.

%TODO these paragraph should contain the summary of evaluation results.
As a proof of concept, we developed \textit{Querypoint}, a prototype
which enhances a popular JavaScript debugger, Firebug, with
\textit{lastChange} feature. Moreover, \textit{Querypoint} provides
mechanisms for automated bug reproduction, and a novel user interface
to investigated execution points and results.

@@

According to \cite{LaToza}, developers spend about fifty percent of
their time debugging. Trial and error, guess-work, and analyzing large
and complicated data make debugging difficult and time-consuming. To fix 
a bug, developers typically reproduce and monitor the buggy execution several 
times to understand the program's unexpected behavior. Therefore, any
enhancement to debugging techniques and tools which automates or facilitates developers' actions
on execution traces, can save developers' time, reduce development costs and improve programs' quality.

@@

To address these problems, we introduce a new functionality in
debuggers, \textit{lastChange}, which locates the origin of a wrong
value. It uses program re-execution similar to the practice
in using breakpoints and log statements. Imagine a program execution 
paused on a breakpoint and the developer is
suspicious about the value of a variable or an object property. The
developer queries \textit{lastChange} on the value. The debugger
reproduces the buggy execution and collects data
and once the execution reaches the same place (i.e., the same
breakpoint hit), it pauses the execution, analyzes the collected data
and shows the location of the last change to the developer. The
developer can also examine the program state at the located point of
execution, and continue debugging by more \textit{lastChange} queries
from that point.

We refer to the breakpoint used in \textit{lastChange} as the
\textit{reproduction point}: this is a point where the developer knows
reproduces the bug and it is the point were the \textit{lastChange}
algorithm ends. The ability to identify this reproduction point, i.e.
 a test case and a breakpoint after the bug has appeared, is the only
prerequisite for \textit{lastChange} functionality.  This means
\textit{lastChange} can be implemented on top of a conventional
breakpoint debugger and it does not require a special environment to
create identical, instruction by instruction, re-executions.

Our contribution in this paper is the technique \textit{lastChange},
which locates the last place a value has changed, gathers other values
from that execution point, and allows \textit{lastChange} operations
from that point. The technique builds on existing breakpoint debugger
technology. We demonstrate the feasibility of the approach with
\textit{Querypoint}, an implementation extending the Firebug
JavaScript debugger. \textit{Querypoint} also provides mechanism for
automated bug reproduction, and a novel user interface which
summarizes investigated execution points and collected results.

@@

@@

By browsing through the code, the developer can quickly determine that the value displayed on the 
button is set at line 22. To start debugging, the developer sets a breakpoint
on line 22. Once the button is clicked, execution is paused at line
22. Figure~\ref{fig:example1} shows the Firebug debugger while the
execution is paused. Firebug has several panels (e.g., HTML, CSS,
Script, DOM, etc.) that each demonstrate one aspect of the Web page.
The Script panel contains the list of all loaded source
files and regular debugging facilities such as setting breakpoints and
stepping. To the right of the script panel, the Watch panel shows the program state
where the developer can examine object and variable values. In our case, the
\texttt{myObject.myProperty} value at the paused point is zero. We expected this value to be \texttt{2}.

To apply backward search strategy for locating defects, the developer
first needs to know the origin of the wrong value. To achieve this
goal using breakpoints, the developer should search code to find all possible places that
\texttt{myObject.myProperty} might get a new value and set breakpoint at these locations. However, an
object and property can be accessed and changed through different
names and methods. There is no simple way to identify these alias or
even their total number.  The developer can make a good guess and set
breakpoints on lines where the property seems to be changed. Then they
re-execute the program and examine the state looking for values that
may lead to the incorrect value observed at line 22. All this work
must be repeated if a new alias is discovered or if some
information related to the buggy result was missed while stopped on
one of the breakpoints.

@@

Looking at line 29, it seems that something is wrong with
\texttt{myCondition.value} which causes line 29 execution. The
developer examines \texttt{myCondition.value} and it is
\texttt{undefined}. The next step is to know when this property got
this value. To do so, the developer reruns \textit{lastChange} command
on \texttt{myCondition.value} at this point. This re-executes the
program and pauses at the reproduction point,  showing line 25-the
place \texttt{oldValue} is assigned to
\texttt{myCondition.value}. To examine \texttt{oldValue}
 at this point, developer enters it to the watch box and after re-execution the value
 is available. Now it is clear that the bug occurs because \texttt{oldValue} is
\texttt{undefined} once execution reaches line 25 (Figure~\ref{fig:example4}).

@@

\begin{figure}[htp]
\includegraphics{5-example-points.jpg}
\caption{The examined points before locating the defect. The arrow represents the logical forward progress of the program. Three actual executions are superimposed on this arrow. All three stop at the reproduction point indicated by the circle 1. After the first execution, the developer asks for lastChange as described in \ref{sec:introExample}, yielding information  indicated by circle 2. After the second execution, another lastChange query causes a third execution, yielding information indicated by circle 3.}
\label{fig:example-points}
\end{figure}

As demonstrated in Figure ~\ref{fig:example-points}, the developer has
examined three points of execution. The first point, the reproduction point,  was the breakpoint set by the developer.
The second and third points preceded the reproduction point in execution sequence.
All three points-the history
of the search for the defect-are available through the debugger's
interface. On the top of the left panel in Figure~\ref{fig:example4}
there is an opened list which shows all three examined points. The
first one is the breakpoint on line 22, the second one is the point
which is when \texttt{myObject.myProperty} changed before
reaching the breakpoint and finally the last one is the point of
execution in which \texttt{myCondition.value} gets the
\texttt{undefined} value. Moreover, the source lines related to these
points are marked with red-cycle icons.

@@

As \textit{lastChange} itself is not a different kind of \textit{breakpoint},
all of the assumptions developers already have for breakpoints hold
for \textit{lastChange}. In particular the re-execution is just the
same operation developers use to debug with breakpoints. Because
\textit{lastChange} dramatically reduces the developers work in
setting and removing breakpoints, the re-execution step will become a
much larger fraction of the debug time. Therefore automatic mechanisms
for re-execution will be corresponding more valuable for debugging
with \textit{lastChange}.

As we described in the preceding section, a \textit{lastChange} query can
be performed on the result of another \textit{lastChange} query. If we
name the reproduction point \textit{R}, we can write the first
\textit{lastChange} in the introductory example in this form:
\textit{lastChange(R, myObject.myProperty)}. It means that this query
is defined at \textit{R}. If we name the result of this query
\textit{A}, we can write the second \textit{lastChange} in this form:
\textit{lastChange(A, myCondition.value)}. In this way, a sequence of
\textit{lastChange} queries with any length can be defined.

\textit{lastChange} can be called on two different types of values: An
object property or a variable value. Three different variations of the
algorithm are required, one for object property and two more depending
upon the scope of a variable.

\subsection{\textit{lastChange} on object property}
To simplify the algorithm explanation and defer technical details, we
define two basic operations and later we explain the details of these
two operations. The first operation is \texttt{objectId()}: given a
JavaScript object it returns an integer as its identifier. This
identifier is unique to the object during one execution.  By using an
object id instead of an object reference we allow the garbage
collector to reclaim the space for dead objects just as it would in the
absence of the debugger. The second
operation is \texttt{setPropertyChangeHook()}: given a function and a
string, the function is called whenever a property changes and its
name matches the string. For example, if the string is 
\texttt{foo} so changes to \texttt{bar.foo} or \texttt{baz.foo} would
call the function.  The callback function receives a reference to the
owner of \texttt{foo}. In the next section, we describe how we use
a slightly different version of this operation which is restricted 
to the created objects in one location instead of all objects.

To see how these functions work, suppose the developer asks for the
last change of \texttt{bar.foo} at the reproduction point in a
program. The debugger calls \texttt{setPropertyChangeHook()} with
\texttt{foo} as the property name and re-executes the
program. Whenever \texttt{foo} changes and the callback function is to
be called, debugger first calls \texttt{objectId()} on the
\texttt{foo} owner object. Then it stores this owner id, the stack
frame locations, and other state values in scope at the call point.
Whenever the execution reaches the reproduction point the debugger
looks at the history of \texttt{foo} changes and finds the last
\texttt{foo} change with the same object id as \texttt{bar} id at the
reproduction point. Figure ~\ref{fig:foo-changes1} shows the list of
property \texttt{foo} change events in a hypothetical
execution. \texttt{bar} id at the reproduction point is 1010, so the
last change of \texttt{bar.foo} is the fourth column.

\begin{figure}[htp]
\includegraphics[width=.48\textwidth]{6-foo-changes1.jpg}
\caption{The list of property \texttt{foo} change events and
  \texttt{bar} id at the reproduction point which identifies the last
  change of \texttt{bar.foo} in the list. Column 3 is also a change of
  the object we want to study, id 1010, but it is not the last change;
  Column 5 is also a change of a property \texttt{foo} but its not for
  the object we are interested in.}
\label{fig:foo-changes1}
\end{figure}

\subsection{\textit{lastChange} on variable} 
In JavaScript, every frame has a scope chain and every available
variable in the frame comes from one of the scopes in the frame's
scope chain. Once a developer asks for the last change of a variable with name 
\texttt{foo} at the reproduction point, the debugger first determines the
variable's scope as follows: it iterates over the scopes in the scope
chain and the first scope which has a variable with the same name is
the variable's scope. There are five different scope types: local,
global, \texttt{with}, closure and \texttt{catch}. We explain these
cases in two groups.

@@

\end{verbatim}
\caption{A recursive call trace demonstrates the need for scope
  id. In the absence of scope id, querying \textit{lastChange} on \texttt{x}
  at point C in scope 1 returns point B which is the last change of variable 
  \texttt{x} but happens in another scope. However, the right answer is point A
  which is in the same scope as point C.}
\label{fig:recursive}
\end{figure}

\subsubsection{global and \texttt{with} scopes}
Global scope is the most outer scope in the scope chain and it is
also referred to as the global object (the \texttt{window}
object in Web pages). This scope is a regular JavaScript object and therefore every
global variable is a property of global object. Similarly,
 \texttt{with} scopes are also regular objects. A \texttt{with}
scope is created by a \texttt{with()} block with an object as the
parameter. Every property of this parameter object is available inside the
block as a variable. \textit{lastChange} treats the case where variable's scope
is global or \texttt{with}, like it does on an object
property.

Having the scope chain and the source code, we can map every scope to % perhaps needs more explanation, how?
a code block, enclosed the executing code. In
JavaScript, a code block can be identified by the file url and  
the block's first instruction program counter. Given this information, the debugger is
able to recognize the code block in loaded scripts or once it is loaded.
Similar to \textit{lastChange} on object property, we define two basic operations: 
\texttt{scopeId()} and \texttt{setVariableChangeHook()}. The first one, 
given a scope returns an integer as the scope's identifier. The second one, 
given a callback function, a code block and a variable name-which is defined inside the block, the function is called
whenever the variable is changed. Figure ~\ref{fig:recursive} demonstrates a case in which the same syntactically variable
\texttt{x} is changed several times but in different scopes due to the
recursive calls. Without considering the scopes the last change of x
at point \texttt{C} in scope 1 is point \texttt{B} which is a change of \texttt{x} in 
scope 2 while the right last change is point \texttt{A}. To resolve this issue, an id
is assigned to the scope by \texttt{ScopeId()} operation and the scope id is also stored 
at every variable change event.

If the developer asks for the last change of variable \texttt{foo} at the 
reproduction point in a program, debugger calls 
\texttt{setVariableChang-} \texttt{eHook()} with the variable's defining
block and name as parameters and re-executes the
program. Whenever \texttt{foo} changes and the callback function is to
be called, debugger first calls \texttt{scopeId()} on the
variable's scope. Then it stores this scope id, the stack
frame locations, and other state values in scope at the call point.
Whenever the execution reaches the reproduction point the debugger
looks at the history of \texttt{foo} changes and finds the last
\texttt{foo} change with the same scope id as the variable's scope id
at the reproduction point (Figure ~\ref{fig:foo-changes2}).

\begin{figure}[htp]
\includegraphics[width=.48\textwidth]{7-foo-changes2.jpg}
\caption{The list of variable \texttt{foo} change events and the
  \texttt{foo}'s scope id at the reproduction point which identifies
  the last change.}
\label{fig:foo-changes2}
\end{figure}

\subsection{\textit{lastChange} on \textit{lastChange}}
The algorithm we explained so far needs an enhancement to support
cascades of \textit{lastChange}s. We first explain the issue
and then the solution. Consider the following points:

\begin{center}
\textit{
 point A : the reproduction point \\
 point B : lastChange(A, bar.x) \\
 point C : lastChange(B, baz.y) 
 }
 \end{center}
According to the algorithm explained in the previous section
the points B and C cannot be identified before reaching point
A. That is, as execution moves through point B we cannot know it will be the point of last change until
we reach A, and we cannot apply the simple algorithm to look for point C until we know B. 
 Therefore, all \texttt{x} change events for point B and all \texttt{y}
change events for point C are stored. Once the execution reaches point A,
and the exact \texttt{bar} id reveals, point B can be identified among
the change events in the history of \texttt{x} changes. The problem
arises in recognizing point C. To recognize point C, we need
\texttt{baz} id at point B, but B is a past program state, and
\texttt{baz} id has not been collected at this point.

To address this issue, we use Algorithm \ref{dependency-analysis}, to
create the list of additional data (object id or scope id) should be collected at a change
event. This algorithm does a dependency analysis between all
\textit{lastChange} queries and computes the list of additional data
required to be collected at each change event. For example in the
mentioned case, \texttt{baz} id-if it is available- should be
collected at all points \texttt{x} changes. Having this additional
data, \texttt{baz} id at point B, it is enough to look at the list
of \texttt{y} changes and find the last change event with the same 
owner id (Figure ~\ref{fig:lastchange-lastchange}).

\begin{algorithm}
\caption{\textit{lastChange} queries dependency analysis.}
\label{dependency-analysis}
\begin{algorithmic}

@@

\end{algorithmic}
\end{algorithm}

\begin{figure}[htp]
\includegraphics[width=.48\textwidth]{8-lastchange-lastchange.jpg}
\caption{The list of change events stored for locating point B, the
  \textit{lastChange} of \texttt{bar.x} at the reproduction point, and
  point C, the \textit{lastChange} of \texttt{baz.y} at point B.}
\label{fig:lastchange-lastchange}
\end{figure}

%---------------------------------------------------------------------------------------------------
\section{JavaScript implementation}
To verify the \textit{lastChange} algorithm we
implemented it in an extension to the Firebug
JavaScript debugger\cite{Firebug}. %\cite{firebug-version-1.6}. 
Firebug itself is an extension of the Firefox browser\cite{Firefox}. %\cite{firefox-version-3.6}
The Firefox JavaScript engine provides a JavaScript debugging interface \cite{JSD} and 
\textit{Querypoint} is developed over this interface. Our prototype implementation defines our four primitive operations in JavaScript code and using techniques which are cumbersome and comparatively slow to execute.  A professionally useful debugger would implement these primitive operations in the Javasript engine.

\subsection{\texttt{objectId()} operation}
\texttt{objectId(obj)} first checks the argument \texttt{obj} for a property \texttt{\_objectId}.
It returns the value if this property is already defined,
otherwise it generates a new id and sets this property. The value of the id is simply an integer incremented 
for each new \texttt{\_objectId} needed. To assign the id to the object we 
set the id as a regular object property. This might change
the program behaviour due to the appearance of this property in 
\texttt{for(property in object)} loops. \texttt{defineProperty} is a standard JavaScript
function which is available in the next version of Firefox\footnote[1]{https://developer.mozilla.org/en/JavaScript/Reference/Global\_Objects\\/Object/defineProperty}. \texttt{defineProperty} receives 
an argument which specifies whether the property is enumerable or not. By setting the enumerable
field \texttt{false} for \texttt{\_objectId}, this property will not
apear in \texttt{for(property in object)} loops and therefore has no
effect on the program execution. Note that \texttt{\_objectId} is
not set for all objects but only those objects that need an id.

@@

For our purposes, the \texttt{watch()} function only covers the case
of global object properties. At the beginning of execution, no object
excepting the global object is available.  For 
the our \textit{lastChange} prototype we created a version of
\texttt{setPropertyChangeHook()}.  The basic strategy is to get a reference to the object just
after its creation, then use \texttt{watch()} function to monitor
property changes in the object. Setting a flag\footnote[2]{DISABLE\_OBJECT\_TRACE defined in jsdIDebuggerService.idl} into the Firefox %\cite{JSD}
JavaScript engine we can get file URL and line number for each object
creation (e.g., myFile.js, line 24). We set a hook on this line
and analyze the source code to determine which object was created.

@@

There is another case for new objects and when they are passed as
arguments to other functions (Figure~\ref{fig:objectCreation}(c)). In these
cases instead of step-over we use step-in and we get the corresponding
argument as a new object.

Although this approach is successful in many ordinary cases, there are some 
cases where this approach doesn't work. For example in a case
where the new object is assigned to \texttt{a[++i]} where i is a variable
(Figure ~\ref{fig:objectCreation}(d)), \texttt{a[++i]} is kept in the list
of assignee statements. Obviously, evaluating this statement doesn't return
a reference to the new object. Our prototype implementation does not handle these kinds of unusual cases yet.

>>>>

\begin{abstract}
Developers often seek the origins of wrong values they see in their
debugger. Their search must be backwards in time: the code causing the
wrong value executed before the wrong value appeared. Searching with
breakpoint- or log- based debuggers demands persistence and
significant experience with the application being debugged. 
We introduce a new, practical feature for debuggers called \textit{lastChange}, 
which automatically locates the last point that a variable or an object property 
has been changed. Starting from a 
program halted on a breakpoint, the \textit{lastChange} solution
applies queries to the live program during re-execution, recording the call stack 
and limited program state each time the property value changes.
When the program halts again on the breakpoint, the recorded information 
can be shown to the developer.
As a proof of this concept, we developed \textit{Querypoint}, a prototype
which enhances the popular Firebug JavaScript debugger with
 the \textit{lastChange} feature and studied users applying the prototype to 
some simple test cases.
 The approach used in implementing 
\textit{lastChange} combines the flexibility of breakpoint debugging
with the expressive power of log-based query debugging.  Contrary to
other replay-based approaches which require exactly the same
re-executions (deterministic executions), our new approach only requires \textit{bug 
reproducibility}, that is a test case is available which reproduces the bug.

@@

According to \cite{LaToza}, developers spend about fifty percent of
their time debugging. To fix a bug, developers typically reproduce 
and monitor the buggy execution several times to understand the 
program's unexpected behavior. Trial-and-error, guess-work, and 
analyzing complicated data make debugging difficult and time-consuming. 
Therefore, enhancements to debugging techniques and tools which automate 
debugging operations can save developer 
time, reducing development costs  and impoving software quality.

@@

To address these problems, we introduce a new functionality in
debuggers, \textit{lastChange}, which locates the origin of a wrong
value. It uses program re-execution similar to the practice
in using breakpoints and manually inserted log statements. 
Imagine a program execution 
paused on a breakpoint and the developer is
suspicious about the value of a variable or an object property. The
developer queries \textit{lastChange} on the value. The debugger
reproduces the buggy execution and collects data
when the data field changes. Once the execution reaches the same place 
(i.e., the same
breakpoint hit), it pauses the execution, analyzes the collected data
and shows the location of the last change to the developer. The
developer can also examine the program state at the located point of
execution, and continue debugging by more \textit{lastChange} queries
from that point.

As we detail in the Related Work section, numerous papers explore forms 
of automatic logging of program state, augmented with queries on the log. 
This replaces the developer time needed to 
insert statement manually with computer time to write the voluminious logs. 
Our approach uses semi-automatic live-execution queries: user interface 
operations allow the developer to select queries and these queries operate on the live program during re-execution.

We refer to the breakpoint used in \textit{lastChange} as the
\textit{reproduction point}: this is any point where the developer knows
the bug has been reproduced. It is also the point were each \textit{lastChange}
algorithm cycle ends. The ability to identify this reproduction point, i.e.
 a test case and a breakpoint after the bug has appeared, is the only
prerequisite for \textit{lastChange} functionality.  This means
\textit{lastChange} can be implemented on top of a conventional
breakpoint debugger and it does not require a special environment to
create identical, instruction by instruction, re-executions.

Our contribution in this paper is the technique \textit{lastChange},
which locates the last place a value has changed, gathers other values
from that execution point, and allows \textit{lastChange} operations
from that point. The technique builds on existing breakpoint debugger
technology. We demonstrate the feasibility of the approach with %TODO
\textit{Querypoint}, an implementation extending Firebug
JavaScript debugger. \textit{Querypoint} also provides mechanism for
automated bug reproduction, and a novel user interface which
summarizes investigated execution points and collected results.

@@

By browsing through the code or other means\cite{Barton2010},
 the developer determines that the value displayed on the 
button is set at line 22. Since the displayed value is incorrect we know 
the bug occured before we hit this line.
To start debugging, the developer sets a breakpoint
on line 22. Once the button is clicked, execution is paused at line
22. Figure~\ref{fig:example1} shows the Firebug debugger while the
execution is paused. Firebug has several panels (e.g., HTML, CSS,
Script, DOM, etc.) that each demonstrate one aspect of the Web page.
The Script panel contains the list of all loaded source
files and regular debugging facilities such as setting breakpoints and
stepping. To the right of the script panel, the Watch panel shows the program state
where the developer can examine object and variable values. In our case, the
\texttt{myObject.myProperty} value at the paused point is zero. We expected this value to be \texttt{2}.

To apply backward search strategy for locating defects, the developer
first needs to know the origin of the wrong value. To achieve this
goal using breakpoints, the developer should search code to find all possible places that
\texttt{myObject. myProperty} might get a new value and set breakpoint at these locations. However, an
object and property can be accessed and changed through different
names and methods. There is no simple way to identify these alias or
even their total number.  The developer can make a good guess and set
breakpoints on lines where the property seems to be changed. Then they
re-execute the program and examine the state looking for values that
may lead to the incorrect value observed at line 22. All this work
must be repeated if a new alias is discovered or if some
information related to the buggy result was missed while stopped on
one of the breakpoints.

@@

@@

Looking at line 29, it seems that something is wrong with
\texttt{myCondition.value} which causes line 29 execution. The
developer examines \texttt{myCondition.value} and it is
\texttt{undefined}. The next step is to know when this property got
this value. To do so, the developer runs the \textit{lastChange} command
on \texttt{myCondition.value} at this point. This re-executes the
program and pauses at the reproduction point,  showing line 25-the
place \texttt{oldValue} is assigned to
\texttt{myCondition.value}. If the developer asks for \textit{lastChange} on \texttt{oldValue}, 
the debugger can notify the developer that this variable is never assigned a value.
 Now it is clear that the bug occurs because \texttt{oldValue} is
\texttt{undefined} once execution reaches line 25 (Figure~\ref{fig:example4}).

As demonstrated in Figure ~\ref{fig:example-points}, the developer has
examined three points of execution. The first point, the reproduction point,  was the breakpoint set by the developer.
The second and third points preceded the reproduction point in execution sequence.
All three points-the history
of the search for the defect-are available through the debugger's
interface. On the top of the left panel in Figure~\ref{fig:example4}
there is an opened list which shows all three examined points. The
first one is the breakpoint on line 22, the second one is the point
which is when \texttt{myObject.myProperty} changed before
reaching the breakpoint and finally the last one is the point of
execution in which \texttt{myCondition.value} gets the
\texttt{undefined} value. Moreover, the source lines related to these
points are marked with red "Q" icons.

@@

\begin{figure}[htp]
\centering

\includegraphics{5-example-points.jpg}
\caption{The examined points before locating the defect. The arrow represents the logical forward progress of the program. Three actual executions are superimposed on this arrow. All three stop at the reproduction point indicated by the circle 1. After the first execution, the developer asks for lastChange as described in \ref{sec:introExample}, yielding information  indicated by circle 2. After the second execution, another lastChange query causes a third execution, yielding information indicated by circle 3.}
\label{fig:example-points}
\end{figure}

@@

As we described in the preceding section, a \textit{lastChange} query can
be performed on the result of another \textit{lastChange} query. If we
name the reproduction point \textit{R}, we can write the first
\textit{lastChange} in the introductory example in this form:
\textit{lastChange(R, myObject.myProperty)}. It means that this query
is defined at \textit{R}. If we name the result of this query
\textit{L}, we can write the second \textit{lastChange} in this form:
\textit{lastChange(L, myCondition.value)}. In this way, a sequence of
\textit{lastChange} queries with any length can be defined.

\textit{lastChange} can be called on object property, on a variable value, 
or on the results of a \textit{lastChange}. We explain each case in the following subsections.

\subsection{\textit{lastChange} on object property}
To simplify the algorithm explanation and defer technical details, we
define two basic operations and later we explain the details of these
two operations. The first operation is \texttt{objectId()}: given a
JavaScript object it returns an integer as its identifier. This
identifier is unique to the object during one execution.  By using an
object id instead of an object reference we allow the garbage
collector to reclaim the space for dead objects just as it would in the
absence of the debugger. The second
operation is \texttt{setPropertyChangeHook()}: given a function and a
string, the function is called whenever a property changes and its
name matches the string. For example, if the string is 
\texttt{foo} changes to \texttt{bar.foo} or \texttt{baz.foo} would
call the function.  The callback function receives a reference to the
owner of \texttt{foo}.

To see how these functions work, suppose the developer asks for the
last change of \texttt{bar.foo} at the reproduction point in a
program. The debugger calls \texttt{setPropertyChangeHook()} with
\texttt{foo} as the property name and re-executes the
program. Whenever \texttt{foo} changes and the callback function is to
be called, debugger first calls \texttt{objectId()} on the
\texttt{foo} owner object. Then it stores this owner id, the stack
frame locations, and other state values in scope at the call point. 
Then the callback returns to continue execution. Thus the query is not 
a breakpoint in the sense of pausing for user interaction, but breakpoint 
technology can be used to implement the query.
Whenever the execution reaches the reproduction point the debugger
looks at the history of \texttt{foo} changes and finds the last
\texttt{foo} change with the same object id as \texttt{bar} id at the
reproduction point. Figure ~\ref{fig:foo-changes1} shows the list of
property \texttt{foo} change events in a hypothetical
execution. \texttt{bar} id at the reproduction point is 1010, so the
last change of \texttt{bar.foo} is the fourth column.

\begin{figure}[htp]
\centering 
\includegraphics[height=.2\textheight]{6-foo-changes1.jpg} 
\caption{A hypothetical list of  change events for a property \texttt{foo}. 
Each change event adds a column with the id of the object changed, the call stack,
and some program state such as local variable values. 
At the reproduction point  we determine which id  corresponds to 
object \texttt{bar} and read out column 4,  the last
  change of \texttt{bar.foo}. Column 3 is also a change of
  the object we want to study, id 1010, but it is not the last change;
  Column 5 is also a change of a property \texttt{foo} but its not for
  the object we are interested in.}
\label{fig:foo-changes1}
\end{figure}

\subsection{\textit{lastChange} on variable} 
In JavaScript, every frame has a scope chain and every available
variable in the frame comes from one of the scopes in the frame's
scope chain. Once the developer asks for the last change of a variable with name 
\texttt{foo} at the reproduction point, the debugger first determines the
variable's scope as follows: it iterates over the scopes in the scope
chain and the first scope which has a variable with the same name is
the variable's scope. There are five different scope types: local,
global, \texttt{with}, closure and \texttt{catch}. We explain these
cases in two groups.

\subsubsection{global and \texttt{with} scopes}
Global scope is the most outer scope in the scope chain and it is
also referred to as the global object (the \texttt{window}
object in Web pages). This scope is a regular JavaScript object and therefore every
global variable is a property of global object. Similarly,
 \texttt{with} scopes are also regular JavaScript objects. A \texttt{with}
scope is created by a \texttt{with()} block with an object as the
parameter. Every property of this parameter object is available inside the
block as a variable. \textit{lastChange} treats the case where variable's scope
is global or \texttt{with}, like it does on an object
property.

Having the scope chain and the source code, we can map every scope to % perhaps needs more explanation, how?
a code block, enclosed the executing code. In
JavaScript, a code block can be identified by the file url and  
the block's first instruction program counter. Given this information, the debugger is
able to recognize the code block in loaded scripts or once it is loaded.
Similar to \textit{lastChange} on object property, we define two basic operations: 
\texttt{scopeId()} and \texttt{setVariableChangeHook()}. The first one, 
given a scope returns an integer as the scope's identifier. The second one, 
given a callback function, a code block and a variable name-which is defined inside the block, the function is called
whenever the variable is changed.

@@

\end{verbatim}
\caption{A recursive call trace illustrating the scope
  id. The lines in the bottom half of the diagram simulate a trace of the change events for the variable \texttt{x} as the function \texttt{f()} calls itself. Each call creates a scope; eventually when the variable \texttt{stop} is changed by an external process we return from the recursion. The lines marked A, B, and C are discussed in the text.}
\label{fig:recursive}
\end{figure}

Figure ~\ref{fig:recursive} illustrates how the \texttt{ScopeId()} operation separates instances 
of a variable in different scopes having the same name.
If we ask for the last change on \texttt{x} at point labeled \texttt{C} in scope with id 1, we want the change at line \texttt{A} in scope 1, not 
the change at the line marked \texttt{B}, where a variable named \texttt{x} in scope 3 is changed.

If the developer asks for the last change of variable \texttt{foo} at the 
reproduction point in a program, debugger calls 
\texttt{setVariableChangeHook()} with the variable's defining
block and name as parameters and re-executes the
program. Whenever \texttt{foo} changes and the callback function is to
be called, debugger first calls \texttt{scopeId()} on the
variable's scope. Then it stores this scope id, the stack
frame locations, and other state values in scope at the call point.
Whenever the execution reaches the reproduction point the debugger
looks at the history of \texttt{foo} changes and finds the last
\texttt{foo} change with the same scope id as the variable's scope id
at the reproduction point (Figure ~\ref{fig:foo-changes2}).

\begin{figure}[htp]
\centering 
\includegraphics[height=.12\textheight]{7-foo-changes2.jpg}
\caption{A hypothetical list of variable \texttt{foo} change events. Each column of the list indicates a change event; for each
change event the scope id returns by \texttt{ScopeId()} is recorded along with call stack and program state information. 
   The last column having the scope id \texttt{foo} at the reproduction point indicates 
  the last change.}
\label{fig:foo-changes2}
\end{figure}

\subsection{\textit{lastChange} on \textit{lastChange}}
The \textit{lastChange} algorithm records changes by id (either object or scope id), then reads out the last change when we
arrive at the reproduction point and discover the id of the requested value.
When we perform \textit{lastChange} based on a previous \textit{lastChange} the query algorithm 
must retain additional information. Consider the following example:
\begin{center}
\textit{
 point A : the reproduction point \\
 point B : lastChange(A, bar.x) \\
 point C : lastChange(B, baz.y) 
 }
 \end{center}
where point A is a breakpoint, point B is the last change of the object property \texttt{bar.x} at point A, and
point C is the last change of the object property \texttt{baz.y} at point B. 
The object referenced by \texttt{baz} changes upon re-execution. Therefore when the developer
asks for the last change of \texttt{baz.y}, we need to track objects named \texttt{baz}
at changes of \texttt{bar.x} and changes to objects named \texttt{y}. Then at the reproduction point
we need to work out which \texttt{baz} the developer wanted, then select the last change of that \texttt{baz.y}.
Figure ~\ref{fig:lastchange-lastchange} illustrates the extra row of data tracking of \texttt{baz} objects 
at \texttt{x} change events 
and how the id values allow the last change of \texttt{baz.y} to be worked out.

In the general case we perform dependency analysis  as outlined in 
Figure~\ref{fig:dependency-analysis} to
create the list of additional data (object id or scope id) to be collected at a change
event. The process can be repeated to cascade \textit{lastChange} arbitrarily deep.

\begin{figure}[htp]
\centering 
%\begin{algorithm}
\begin{algorithmic}

@@

\end{algorithmic}
%\end{algorithm}
\caption{\textit{lastChange} queries dependency analysis.}
\label{fig:dependency-analysis}
\end{figure}

\begin{figure}[htp]
\centering 
\includegraphics[height=.3\textheight]{8-lastchange-lastchange.jpg}
\caption{The list of change events stored for locating point B, the
  \textit{lastChange} of \texttt{bar.x} at the reproduction point, and
  point C, the \textit{lastChange} of \texttt{baz.y} at point B.}
\label{fig:lastchange-lastchange}
\end{figure}

%---------------------------------------------------------------------------------------------------
\section{JavaScript implementation}
To verify the \textit{lastChange} algorithm we
implemented it in an extension to the Firebug
JavaScript debugger\cite{Firebug}. %\cite{firebug-version-1.6}. 
Firebug itself is an extension of the Firefox browser\cite{Firefox}. %\cite{firefox-version-3.6}
The Firefox JavaScript engine provides a JavaScript debugging interface \cite{JSD} and 
\textit{Querypoint} is developed over this interface. Our prototype implementation defines our four primitive operations in JavaScript code and using techniques which are cumbersome and comparatively slow to execute. However the JavaScript prototype is
easy to explore, change, and share with others for feedback.
  A professionally useful debugger would implement these primitive operations within the JavaScript engine.

\subsection{\texttt{objectId()} operation}
\texttt{objectId(obj)} first checks the argument \texttt{obj} for a property \texttt{\_objectId}.
It returns the value if this property is already defined,
otherwise it generates a new id and sets this property. The value of the id is simply an integer incremented 
for each new \texttt{\_objectId} needed. This use of a regular object property  might change
the program behaviour. For example it will apear as an extra property in 
\texttt{for(property in object)} loops. 
The next version of Firefox\footnote[1]{https://developer.mozilla.org/en/JavaScript/Reference/Global\_Objects\\/Object/defineProperty}.
provides an alternative via \texttt{defineProperty} a standard JavaScript
function which receives 
an argument specifying whether the property is enumerable or not. By setting the enumerable
field \texttt{false} for \texttt{\_objectId}, this property will not
apear in \texttt{for(property in object)} loops and therefore has no
effect on the program execution. Note that \texttt{\_objectId} is
not set for all objects but only those objects that need an id.

@@

For our purposes, the \texttt{watch()} function only covers the case
of global object properties. At the beginning of execution, no object
excepting the global object and its predefined properties is available.  For 
our \textit{lastChange} prototype we created a version of
\texttt{setPropertyChangeHook()}.  The basic strategy is to get a reference to the object just
after its creation, then use \texttt{watch()} function to monitor
property changes in the object. Setting a flag\footnote[2]{DISABLE\_OBJECT\_TRACE defined in jsdIDebuggerService.idl} into the Firefox 
JavaScript engine, we can get file URL and line number for each object
creation (e.g., myFile.js, line 24). We set a hook on this line
and parse the source code to determine which object was created.

@@

There is another case for new objects where they are passed as
arguments to other functions (Figure~\ref{fig:objectCreation}(c)). In these
cases instead of step-over we use step-in and we get the corresponding
argument as a new object.

Although this approach is successful in many ordinary cases, there are some 
cases that a more comprehensive analysis needed for correct behaviour. For example in a case
where the new object is assigned to \texttt{a[++i]} where i is a variable
(Figure ~\ref{fig:objectCreation}(d)), \texttt{a[++i]} is kept in the list
of assignee statements. Obviously, evaluating this statement doesn't return
a reference to the new object. Our prototype implementation does not handle these kinds of unusual cases yet.

Throughout this section we implicitly assumed that the object will be created at 
the same location in the next execution. So, what does happen if it
is not true? Well, once the execution reaches the reproduction point, it
reveals that the object has been created in different location. This time,
prototype re-executes the program considering both locations as possible 
object creation locations.
